Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-21 probabilisticProgrammingMartingales.tex
Fix geometric distribution example

\lo{NOTATION. To be consistent, we write $\lambda x. M$ rather than $\lambda x : M$.}

\begin{example}[Geometric distribution]
Consider the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \, \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{enumerate}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
\end{enumerate}
\akr{TODO: make this into a diagram with arrows.}

\lo{Here is a more readable version.}

Writing $\Theta = \lambda f, n : \tif{\tsample - 0.5 < 0}{n}{f (n+1)}$, we have
\begin{enumerate}
    \item $(\tY \, \Theta) \; \underline{0}$

    \item $\big(\lambda z.\Theta \; (\tY \, \Theta) \; z \big) \; \underline{i}$

    \item $\Theta \; (\tY \, \Theta) \; \underline{i}$

    \item \changed[akr]{$(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \; \underline{i}$}

    \lo{I think the preceding should be: $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (n+1)}) \; \underline{i}$, and so, more succinctly:
    \[
    \big(
    \lambda n: 
    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
    \big) 
    \; \underline{i}
    \]}

    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{
    (\tY \, \Theta) \, (\underline{i}+1)}$

    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$

    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \, \Theta) (\underline{i}+1)}$

    \item $\underline{i}$

    \item $(\tY \, \Theta) \; (\underline{i}+1)$

    \item $\big(\lambda z. \Theta \; (\tY \, \Theta) \;  z\big) \; (\underline{i} + 1)$

    \item $\Big(\lambda z. \big(
    \lambda n: 
    \tif{\tsample - 0.5 < 0}{n}{(\tY \, \Theta) \; (n+1)} 
    \big) \, z \Big)
    \; \underline{i}$
\end{enumerate}

\lo{So we have:
\[
\begin{array}{ll}
1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
7 \to 9 \to 11 \to \cdots
\end{array}
\]
You have 2, 3 and 10 because your $\tY$-redex rule is
\[
\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
\]
which is not the official definition.}
\end{example}

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
(We will explain why later.)

@@

\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n. n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a partial ranking function:
\begin{align*}
(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
\underline i \mapsto 0
\end{align*}

\section{Antitone ranking functions}

>>>>

\begin{example}[Geometric distribution]
Consider the term
\begin{align*}
&(\tY \lambda f, n. \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \, \underline{0}
\end{align*}
which generates a geometric distribution (let $\Theta = \lambda f, n : \tif{\tsample \underline{- 0.5} < 0}{n}{f (n\underline{+ 1})}$, then the term is $(\tY \Theta)\ \underline 0$).
    Despite its simplicity, its $Rch$ contains all the terms

\begin{tikzpicture}[node distance=0.5em, anchor=west]
    \node (1) at (0,-0) {$\tY\ \Theta\ \underline 0$};
    \node (2) at (0,-1) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ \underline i$};
    \node (3) at (0,-2) {$(\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})})\ \underline i$};
    \node (4) at (0,-3) {$\tif{\tsample \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
    \node (5) at (0,-4) {$\tif{\underline r\ \underline{- 0.5} < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
    \node (6) at (0,-5) {$\tif{\underline r < 0}{\underline i}{\tY\ \Theta\ (\underline i\ \underline{+ 1})}$};
    \node (7) at (2,-6) [text width=] {$\underline i$};
    \node (8) at (0,-7) {$\tY\ \Theta\ (\underline i\ \underline{+ 1})$};
    \node (9) at (0,-8) {$(\lambda z. (\lambda n. \tif{\tsample \underline{- 0.5} < 0}{n}{\tY\ \Theta\ (n \underline{+ 1})}) z)\ (\underline i\ \underline{+ 1})$};

    \draw [->] (1.south west) to [bend right=45] (2.north west);
    \draw [->] (2.south west) to [bend right=45] (3.north west);
    \draw [->] (3.south west) to [bend right=45] (4.north west);
    \draw [->] (4.south west) to [bend right=45] (5.north west);
    \draw [->] (5.south west) to [bend right=45] (6.north west);
    \draw [->] (6.south west) to [bend right=45] (8.north west);
    \draw [->] (8.south west) to [bend right=45] (9.north west);

    \draw [->] (9.west) to [bend left=45] (2.west);
    \draw [->] (6) to (7);
\end{tikzpicture}

Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as having the same value as the next term, so that overall it takes only 3 distinct values.
(We will explain why later.)

@@

\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY \lambda f n.\ n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a partial ranking function:
\begin{align*}
(\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ N & \mapsto 2 \\
\underline i \oplus_{0.5} (\tY \lambda f n.\ n \oplus_{0.5} f(n+1))\ (\underline i + 1) & \mapsto 1 \\
\underline i & \mapsto 0.
\end{align*}

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
\section{Antitone ranking functions}
